\myTitle

### 2015-02-11 tex/main-body-text.tex
edited conclusion

\begin{figure*}
\centering
\begin{minipage}{.45\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{images/Figures-Pat/FlockvFlocktureLoss.pdf}
  \captionof{figure}{Zero-one loss function for {\sc flock} and {\sc flockture} for 9 replicate runs of simulated datasets comprised 
of five populations with varying levels of population differentiation:  $F_{ST} = $ (a) 0.03, (b) 0.02, (c) 0.01, and (d) 0.005.}
  \label{fig:FvFloss}
\end{minipage}%
\hfill
\begin{minipage}{.45\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{images/Figures-Pat/FlockvFlockture_qi.pdf}
  \captionof{figure}{Normalized likelihood values for {\sc flock} and {\sc flockture} for a single run of dataset 1 
with average pairwise $F_{ST} = $ 0.03. }
  \label{fig:FvFqi}
\end{minipage}
\end{figure*}

@@

\subsection*{{\sc flockture} and {\sc structure}} 
As expected {\sc flockture} and {\sc structure} both perform extremely well on datasets with
well differentiated populations. Infrequently, at higher levels of population differentiation the 
NA-NC and NA-C models in {\sc structure} could be trapped in a local mode resulting in the 
assignment of individuals from two groups into one while individuals from another 
were separated into two populations (Fig.~\ref{fig:FvSdistruct}  Simulation 1). This phenomena 
(only present in the microsatellite datasets) disappears as population differentiation 
declines. As the subtlety of genetic structure 
increased loss for both programs increases and differences among markers emerged.
For the microsatellite dataset we observed that both programs
produce roughly equivalent results at higher levels of population differentiation. At intermediate 
values, the variance of the loss incurred by each model increases and certain models do marginally
better in assignment. No model did consistently better than others, until $F_{ST}$ drops 
below 0.0136 after which the loss for {\sc flockture} was markedly larger as was the among run
variance. The drastic increase in variance among runs was also observed with the SNP datasets. 
 {\sc flockture} did, however, have consistenlty smaller loss than  {\sc structure} at high to intermediate 
values of $F_{ST}$. Once $F_{ST}$ fell below 0.016 {\sc structure} began correctly assigning a 
higher proportion of individuals.

The number of plateaus {\sc flockture} encountered and the number of runs that ended on the same plateau 
diminished as the amount of differentiation among populations decreased. For the microsatellite datasets
only one run at $F_{ST} = 0.0339$ produced a single plateau of length greater than six. No plateau record had
a length of six or greater at any $F_{ST}$ value greater than 0.0337. Results from the SNP datasets 
were concordant. No runs of {\sc flockture} resulted in a single plateau length of greater than six and
only $F_{ST}$ values greater 0.0323 had plateau records with a plateau greater than six. Despite the lack of 
plateaus observed for many of the simulated datasets graphs of the $q_i$ values were suggestive 
of population structure within the 
data (Fig.~\ref{fig:FvSdistruct}  Simulation 13).

  \begin{figure*}
\centering
\includegraphics[width=.9\linewidth]{images/Figures-Pat/uSatloss.pdf}%Need to make a greyscale version
  \caption{Zero-one loss over 9 replicate runs of the admixture and correlated allele 
  frequency (A.C) model, 
the no-admixture and correlated allele frequency (NA.C) model, 
and the no-admixture and non-correlated allele frequency (NA.NC) model in {\sc structure} and {\sc flockture}.}
  \label{fig:uSatLoss}
\end{figure*}

 \begin{figure*}
\centering
  \includegraphics[width=.9\linewidth]{images/Figures-Pat/SNPloss.pdf}%Need to make a greyscale version
  \caption{Zero-one loss over 9 replicate runs of the admixture and correlated allele 
  frequency (A.C) model, 
the no-admixture and correlated allele frequency (NA.C) model, 
and the no-admixture and non-correlated allele frequency (NA.NC) model in {\sc structure} and {\sc flockture}.}
  \label{fig:SNPloss}
\end{figure*}

@@

% haven't touched these yet...

\section*{Conclusions}
In the past decade, the use of genetic markers to identify population structure and to attempt inference of
the number of genetic clusters ($K$) in a collection has dramatically increased in the fields of ecology, evolution, 
epidemiology, and conservation
genetics. The most commonly used program for doing this is {\sc structure} \citep{Pritchardetal2000,Falushetal2003}, 
a full-featured program with many options that is widely-used, well-tested, and universally regarded as a standard tool amongst molecular ecologists. 
We've shown here that the model incorporated by {\sc flock} is essentially equivalent to the no-admixture model 
with non-correlated allele frequencies within the program {\sc structure}. In comparing results from {\sc flock} and {\sc structure} 
on simulated datasets with varying levels of differnetiation among populations
we find that the two programs give similar results.

Though we observed similar results
between the two programs at low values of $K$, {\sc flock}'s `best-run' solutions became somewhat inconsistent at higher values of $K$.
This observation was consistent with \citet{Duc&Tur2012}'s comparison of the 
two programs using simulated data 
for $K=4$ \& 8, $L=10$ (Fig. 4 in \citet{Duc&Tur2012}). 
This is likely a consequence of the rapidly increasing size of the search space over possible 
partitions as $K$ and $n$, the number of individuals, increase. 
The likelihood surface defined on partitions of this space will often
be multimodal---possessing numerous local peaks and troughs.
The peak that  {\sc flock}'s algorithm finds is entirely determined by the initial, random, allocation of 
individuals, as the algorithm proceeds deterministically after the initial random allocation.
If the number of initial allocations does not increase
as $K$ is increased, the starting allocations become diffusely scattered in the space of all
possible partitions making it less likely that many (if any at all) will converge to the same local mode. 
This coupled with the fact that there is 
ambiguity in what defines a `best run' when all plateaus are of length zero complicates the interpretation
of results from {\sc flock} if no non-zero plateaus are encountered. The problem of getting 
caught in local modes is not unique to {\sc flock}---{\sc structure} also can converge to 
different solutions in complex spaces; however our results suggest this is a bigger problem for {\sc flock}.
By increasing the number of loci used in their simulations, \citet{Duc&Tur2012} showed  that 
{\sc flock} was able to converge in a simulated data set with $K=8$, but held no advantage in inference of $K$ 
over {\sc structure}. That simulated data set, with $n=240$ was considerably smaller and less complex than the real data investigated here.
%They simulated data using 500 individuals per population but only used 30 of them in the analysis.
% if you have more loci than individuals per population with high Fst between groups it does great!

One of the advantages of {\sc flock} over other methods for genetic clustering
is reported to be the short processing time for each run of {\sc flock}.
While \citet{Duc&Tur2012} observed faster processing times (relative to other methods) in their 
comparisons, we did not observe this advantage in our analysis. This appears to result from two issues. First,
our example data set was an order of magnitude larger than the largest data set analyzed in \citet{Duc&Tur2012}.  
Second, in the comparison by \citet{Duc&Tur2012}, 
{\sc flock} was set to perform 50 runs and 20 iterations for each $K$ starting at 2 and ending when 
one of the stopping conditions was met, and computing times for that analysis were compared to ten iterations of 
the admixture and correlated allele frequency model in {\sc structure} using 
a 50,000 sweep burn-in followed 
by a 200,000 sweep sample from the posterior for values of $K$ from 1 to three more than the true 
value of $K$. 
To us it seemed reasonable to compare, for each $K$, the running times of one iteration of {\sc structure} (50,000 
burn in and 200,000 sample collection sweeps)
to one iteration of {\sc flock} (50 different starting allocations that are reallocated 20 times) 
as both were used to provide one of 6 total estimates (per value of $K$) of the partition of individuals into 
clusters. Though
it is difficult to reliably assess differences in program run times when there is no clear consensus on what 
constitutes a reasonable comparison with respect to run length and number of runs, our results indicate 
that {\sc flock} did not carry a distinct run-time advantage over {\sc structure}.

While there has been lengthy 
discourse over the choice of estimators for $K$ in {\sc structure} 
\citep{Pritchardetal2000,Evannoetal2005,Wap&Gag2006,Gaoetal2011}, less has been written on how the number of 
individuals
and the number of starting allocations can affect not only the convergence of different runs of {\sc flock}, but also its 
inference of $K$. \citeauthor{Duc&Tur2012} developed estimation rules for $K$ based on simulated data 
sets with a modest number of individuals and different migration regimes and rates, numbers of loci, and $K$. 
By relying on the number of identical partitions (mean LLOD values) obtained by {\sc flock} over multiple
runs as a means of supporting 
a particular $K$,  similar---but not identical---partitions
will provide no support to a given $K$. Given {\sc flock}'s tendency to converge to 
different solutions with large $K$ and $n$, it seems that 
this method for estimating $K$ is unlikely to be reliable or workable in large, complex
problems.  One possible patch for this problem would be running
multiple small batches and using the programs {\sc clumpp} \citep{Jak&Ros2007} and
{\sc distruct} \citep{Rosenberg2004} to visually inspect the `best run' results, and then 
determine plateaus amongst {\sc flock}'s replicate solutions not by identical LLOD, but by visual inspection of 
partitions that
are very similar but are perhaps not identical; however such an approach would be labor-intensive
and would rely on arbitrary, difficult-to-reproduce decisions.  It is probably better to use {\sc structure} in such cases, keeping in mind, of course, that
estimating $K$ is a hard problem and should only be undertaken with caution, regardless of the method 
used. This is especially true in real populations which do not conform to the assumptions of genetic clustering models.

We made our comparison between {\sc flock} and the {\em no-admixture} model in 
{\sc structure} which is not the default model used in {\sc structure}. Rather, by default, {\sc structure} uses 
its {\em with-admixture} model.  Such a model explicitly tries to account for the admixed origin
of individuals in the sample and thus may be a more appropriate model than the no-admixture model,
or {\sc flock}, when the sample contains individuals that are admixed between populations.  A further advantage
of the with-admixture model is that it allows inference of the population of origin of individual alleles at 
different loci within individuals, and not just of the individuals themselves.   We also note that {\sc flock} 
corresponds to {\sc structure}'s model with uncorrelated allele frequency priors. \citet{Falushetal2003} found that
using their ``$F$-model,'' which assumes that allele frequencies between populations are correlated {\em a priori}, 
allowed {\sc structure} to dissect more subtle population structure than was possible using the prior with
uncorrelated allele frequencies.  Modifying {\sc flock} so as to reap similar benefits from using a prior
with correlated allele frequencies seems like it would be an interesting challenge, though difficult
because {\sc flock}'s formulation effectively integrates the allele frequencies out of the model.

@@

\section*{Acknowledgments}
We are grateful to Kristen Ruegg and Robin Waples for helpful comments on a draft of this paper.

>>>>

\begin{figure*}
\centering
\begin{minipage}{.45\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{images/Figures-Pat/FlockvFlocktureLoss.pdf}
  \captionof{figure}{Zero-one loss function for {\sc flock} and {\sc flockture} for 9 replicate runs of simulated datasets comprised 
of five populations with varying levels of population differentiation:  $F_{ST} = $ (a) 0.03, (b) 0.02, (c) 0.01, and (d) 0.005.}
  \label{fig:FvFloss}
\end{minipage}%
\hfill
\begin{minipage}{.45\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{images/Figures-Pat/FlockvFlockture_qi.pdf}
  \captionof{figure}{Normalized likelihood values for {\sc flock} and {\sc flockture} for a single run of simulated
data with average pairwise $F_{ST} = $ 0.03. }
  \label{fig:FvFqi}
\end{minipage}
\end{figure*}

@@

\subsection*{{\sc flockture} and {\sc structure}} 
As expected {\sc flockture} and {\sc structure} both perform extremely well on datasets with
well differentiated populations. Infrequently, at higher levels of population differentiation the 
NA-NC and NA-C models in {\sc structure} exhibited large losses (Fig.~\ref{fig:uSatLoss})
resulting from the assignment of individuals from two groups into a single cluster while 
individuals from another group were separated into two clusters
with intermediate $q_i$ values (Fig.~\ref{fig:FvSdistruct}  A). This phenomena 
(only present in a single microsatellite dataset) disappears as population differentiation 
declines. As the subtlety of genetic structure 
increased loss for both programs increases and differences in the performance of
each model with different marker types emerged.
For the microsatellite dataset we observed that both programs
produce roughly equivalent results at higher levels of population differentiation. At intermediate 
values, the variance of the loss incurred by each model increases and certain models do marginally
better in assignment. No model did consistently better than others, until $F_{ST}$ drops 
below 0.0136 after which the loss for {\sc flockture} was markedly larger as was the among run
variance. The drastic increase in variance among runs was also observed with the SNP datasets. 
 {\sc flockture} did, however, have consistenlty smaller loss than  {\sc structure} at high to intermediate 
values of $F_{ST}$. Once $F_{ST}$ fell below 0.016 {\sc structure} began correctly assigning a 
higher proportion of individuals.

The number and length of plateaus {\sc flockture} encountered quickly 
diminished as the amount of differentiation among populations decreased. For the microsatellite datasets
only one run at $F_{ST} = 0.0339$ produced a single plateau of length greater than six. The other eight 
replicate runs yielded plateau records varying in length from two to eight all of which had at 
least one plateau larger than thirteen. When $F_{ST}$ was larger than 0.0337 no plateau record
contained at least one plateau of length six or greater. No plateaus were observed for any run when 
$F_{ST}$ was larger than 0.02187. Results from the SNP datasets
were concordant. No run of {\sc flockture} resulted in a single plateau of length greater than six.
Only simulated datasets with pairwise $F_{ST}$ values greater 0.0323 had plateau records with a 
plateau greater than six. Despite the lack of plateaus observed for many of the simulated datasets 
graphs of the normalized likelihood values from the {\sc flockture} run with the highest 
likelihood were suggestive of population structure.

Plots of the normalized likelihood values from {\sc flockture} can be useful in discerning the presence of 
structure.  In situations where only a single run of {\sc flockture} produced a plateau of length 2 
(Fig.~\ref{fig:FvSdistruct} B) the variation in assignment of individuals among runs was indistinguisable
and results were highly concordant with those of {\sc structure}. As population differentiation degrades 
and plateaus are not observed, the run with the highest likelihood from {\sc flockture} still yields
similar clusters as {\sc structure} but with more variation among runs (Fig.~\ref{fig:FvSdistruct} C \& D).
As pairwise $F_{ST}$ decreases and the variation among {\sc flockture} runs increases the normalized likelihood
values for assignment remain high in contrast to the $q_i$ values in each of the three models of {\sc structure}.

  \begin{figure*}
\centering
\includegraphics[width=.9\linewidth]{images/Figures-Pat/uSatloss.pdf}%Need to make a greyscale version
  \caption{Zero-one loss for nine runs of {\sc flockture} and the admixture and correlated allele 
  frequency (A.C) model, the no-admixture and correlated allele frequency (NA.C) model, 
and the no-admixture and non-correlated allele frequency (NA.NC) model in {\sc structure}. Forty-eight 
datasets were simulated with varying levels of population differentation. A coalescent based program
was used to generate 15 loci microsatellite genotypes for 2000 individuals comprising five populations.}
  \label{fig:uSatLoss}
\end{figure*}

 \begin{figure*}
\centering
  \includegraphics[width=.9\linewidth]{images/Figures-Pat/SNPloss.pdf}%Need to make a greyscale version
  \caption{Zero-one loss for nine runs of {\sc flockture} and the admixture and correlated allele 
  frequency (A.C) model, the no-admixture and correlated allele frequency (NA.C) model, 
and the no-admixture and non-correlated allele frequency (NA.NC) model in {\sc structure}. Forty-eight 
datasets were simulated with varying levels of population differentation. A coalescent based program
was used to generate 96 loci single nucleotide polymorphism (SNP) genotypes for 2000 
individuals comprising five populations.}
  \label{fig:SNPloss}
\end{figure*}

@@

\section*{Conclusions}
In the past decade, the use of genetic markers to identify population structure and to attempt inference of
the number of genetic clusters ($K$) in a collection has dramatically increased in the fields of ecology, evolution, 
epidemiology, and conservation
genetics. The most commonly used program for doing this is {\sc structure} \citep{Pritchardetal2000,Falushetal2003}, 
a full-featured program with many options that is widely-used, well-tested, and universally regarded as a standard tool amongst molecular ecologists. 
Estimating $K$ is a difficult task, and because it is often based on individual assignment we 
evaluated these programs on their ability to correctly assign indivdiuals at the true value
of $K$. We have shown here that the model incorporated by {\sc flock} is essentially equivalent to the no-admixture model 
with non-correlated allele frequencies within the program {\sc structure}. 
We adapted the source code of {\sc structure} to create {\sc flockture} and show that
 {\sc flock} and {\sc flockture} are essentially equivalent programs. 
In comparing results from {\sc flockture} and {\sc structure} 
on simulated datasets with varying levels of differnetiation among populations
we find that the two programs give similar results.

Though we observed similar results
between the two programs at large values of $F_{ST}$, 
{\sc flockture}'s `best-run' solutions consistently assigned fewer individuals correctly
and became somewhat inconsistent at lower values of $F_{ST}$.
This is likely a consequence of the low population divergence and also
the large size of the search space over possible partitions. 
The likelihood surface defined on partitions of this space will often
be multimodal---possessing numerous local peaks and troughs.
The peak that  {\sc flockture}'s algorithm finds is entirely determined by the initial, random, allocation of 
individuals, as the algorithm proceeds deterministically after the initial random allocation.
By having a large number of individuals
that can be arranged into 5 groups defines a large partion space.
Decreasing $F_{ST}$ leads to more homogenous 
populations whose multilocus genotypes have a smaller difference in 
likelihood of assigning to any given cluster. This could act by increasing the number of
local peaks, leading {\sc flockture} settle on many different solutions. The problem of getting 
caught in local modes is not unique to {\sc flock}---{\sc structure} also can converge to 
different solutions in complex spaces; however our results suggest this is a bigger problem for {\sc flockture}.

One of the advantages of {\sc flock} over other methods for genetic clustering
is reported to be the short processing time for each run of {\sc flock}.
While \citet{Duc&Tur2012} observed faster processing times (relative to other methods), 
their comparision was based on setting {\sc flock} to perform 50 runs and 20 iterations for each 
$K$ starting at 2 and ending when one of the stopping conditions was met, and computing times for that analysis 
were compared to ten iterations of the admixture and correlated allele frequency model in {\sc structure} using 
a 50,000 sweep burn-in followed by a 200,000 sweep sample 
from the posterior for values of $K$ from 1 to three more than the true 
value of $K$. It is difficult to reliably assess differences in program run times when there is no clear consensus on what 
constitutes a reasonable comparison with respect to run length and number of runs, however, we show that if 
{\sc flock} were to be implemented in a coding language rather than a VBA routine in microsoft excel
it would be approximately 200 times faster.

While there has been lengthy 
discourse over the choice of estimators for $K$ in {\sc structure} 
\citep{Pritchardetal2000,Evannoetal2005,Wap&Gag2006,Gaoetal2011}, less has been written on how the number of 
individuals
and the number of starting allocations can affect not only the convergence of different runs of {\sc flock}, but also its 
inference of $K$. \citeauthor{Duc&Tur2012} developed estimation rules for $K$ based on simulated data 
sets with a modest number of individuals and different migration regimes and rates, numbers of loci, and $K$. 
By relying on the number of identical partitions (mean LLOD values) obtained by {\sc flock} over multiple
runs as a means of supporting 
a particular $K$,  similar---but not identical---partitions
will provide no support to a given $K$. We have shown that even at intermediate
values of $F_{ST}$ {\sc flockture} has a tendency to converge to 
different solutions where no plateaus are encountered. Despite the lack of
plateaus plots of the normalized likelihood values demonstrate that
{\sc flockture} still produces valuable information. One possible patch for this problem would be running
multiple small batches and using the programs {\sc clumpp} \citep{Jak&Ros2007} and
{\sc distruct} \citep{Rosenberg2004} to visually inspect the `best run' results, and then 
determine plateaus amongst {\sc flock}'s replicate solutions not by identical LLOD, but by visual inspection of 
partitions that
are very similar but are perhaps not identical; however such an approach would be labor-intensive
and would rely on arbitrary, difficult-to-reproduce decisions.  It is probably better to use {\sc structure} 
in such cases, keeping in mind, of course, that
estimating $K$ is a hard problem and should only be undertaken with caution, regardless of the method 
used. This is especially true in real populations which do not conform to the assumptions of genetic clustering models.

@@

\section*{Acknowledgments}
We are grateful to Kristen Ruegg and Robin Waples for helpful comments on a draft of this paper.
P. Barry was supported by a fellowship from the Rasmuson Foundation.
